---
title: "p8105_hw3_yl5829"
output: github_document
date: "2024-10-12"
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)

library(p8105.datasets)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

##Problem 1

```{r}
#load the data
data("ny_noaa")

#data clean
ny_noaa_clean <- ny_noaa %>%
  separate(date, into = c("year", "month", "day"), convert = TRUE) %>%
  mutate(
    tmax = as.numeric(tmax),  
    tmin = as.numeric(tmin),
    snowfall = as.numeric(snow)    
  ) %>%
  filter(!is.na(tmax) & !is.na(tmin))  
print(ny_noaa_clean)
```




##Problem 2

```{r}
demo <- read.csv("data/nhanes_covar.csv")
accel <- read.csv("data/nhanes_accel.csv")

#tidy demographic data by renaming columns
demo <- demo %>%
  rename(
    seqn = X,
    sex = X1...male,
    age = X.1,
    BMI = X.2,
    education = X1...Less.than.high.school 
  ) %>%
  mutate(
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
    education = factor(education, levels = c(1, 2, 3), labels = c("Less than High School", "High School", "More than High School"))
  )

accel <- accel %>% rename(seqn = SEQN)

# Merge two data
merged_data <- merge(demo, accel, by = "seqn")

merged_data <- merged_data %>% 
  mutate(age = as.numeric(age)) 

merged_data <- merged_data%>%
  filter(age >= 21) %>%
  filter(!is.na(sex) & !is.na(age) & !is.na(BMI) & !is.na(education)) 

print(merged_data)

```

```{r}
#summary table for the number of men and women in each education category
education_sex_table <- merged_data %>%
  group_by(education, sex) %>%
  summarize(count = n()) %>%
  arrange(education, sex)

print(education_sex_table)

```
By looking at the table we can see that both male and female show higher education level, and there is no large gap between education level of male and female.


```{r}
# Create a line plot for age distribution by education level and sex
ggplot(merged_data, aes(x = age, color = sex, group = sex)) +
  geom_density() +  # Line plot using density for smoother distribution
  facet_wrap(~ education) +
  labs(title = "Age Distribution by Education Level and Sex", 
       x = "Age", 
       y = "Density", 
       color = "Sex") +
  theme_minimal()

```
By looking at the plot we see that there are differences in the ages of participants based on both education level and sex. Female tend to peak at younger ages in each education category, while male age distributions are generally broader.

```{r}
#aggregate total activity
activity <- merged_data %>%
  rowwise() %>%
  mutate(total = sum(c_across(starts_with("min")), na.rm = TRUE)) %>%
  ungroup()

print(activity)

```

```{r}
#plot
ggplot(activity, aes(x = age, y = total, color = sex)) +
  geom_point(alpha = 0.6) +  
  geom_smooth(method = "loess", se = FALSE) +  
  facet_wrap(~ education) +  
  labs(title = "Total Activity vs Age by Education Level and Sex", 
       x = "Age", 
       y = "Total Activity", 
       color = "Sex") +
  theme_minimal()

```

In education level equal or higher than high school, female shows more activity levels than male.
In Less than high school category, both male and female shows a suddenly increase activity level around age 60. However, overall both male and female illustrate a obviously decline trend. This trend is also obvious for more than high school group. 
There is a big upward trending from age 21 to age 40 for high school group, which is significant different compare to any other groups.

```{r}
#make the data from wide to long format for the minute columns
long_activity_data <- merged_data %>%
  pivot_longer(cols = starts_with("min"), names_to = "minute", values_to = "activity") %>%
  mutate(minute = as.numeric(gsub("min", "", minute)))

#plot
ggplot(long_activity_data, aes(x = minute, y = activity, color = sex)) +
  geom_line(alpha = 0.3) +  
  geom_smooth(se = FALSE, method = "loess") +  
  facet_wrap(~ education) +
  labs(title = " 24-hour activity time courses", 
       x = "Minute of the Day", 
       y = "Activity Level", 
       color = "Sex") +
  theme_minimal()

```

By observing the plot we see that activity generally increases in the morning, reach its peaks around midday, and declines in the evening across all education levels. People has "More than High School" education shows higher and more sustained activity throughout the day compared to the other groups. For all education groups, Females consistently show slightly higher activity levels than males, particularly in the middle of the day. In the "Less than High School" group, activity is generally lower, while those in the "High School" and "More than High School" groups maintain more structured and consistent activity patterns. Overall, both education and sex influence the intensity and distribution of physical activity throughout the day.

##Problem3

```{r}
july24 <- read.csv("data/July 2024 Citi.csv") %>%
  mutate(year = 2024, month = "July")
jan24 <- read.csv("data/Jan 2024 Citi.csv") %>%
  mutate(year = 2024, month = "January")
july20 <- read.csv("data/July 2020 Citi.csv") %>%
  mutate(year = 2020, month = "July")
jan20 <- read.csv("data/Jan 2020 Citi.csv") %>%
  mutate(year = 2020, month = "January")

#combine all datasets
citi <- bind_rows(july24, jan24, july20, jan20)

#clean the combined data
citi <- citi %>%
  drop_na() %>%
  mutate(
    weekdays = factor(weekdays, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")),
    duration = as.numeric(duration), 
    member_casual = as.factor(member_casual),
    rideable_type = as.factor(rideable_type)
  )

print(citi)
```
The dataset contains around 99253 rows, each representing a ride on the Citi Bike system.
Each row has information on the ride type (electric or classic bike), the weekdays the ride was taken, the duration (in minutes), the start station name, the end station name, and whether the rider was a member or a casual rider.

```{r}
#make a table of total rides by year, month, and membership status
summary <- citi %>%
  group_by(year, month, member_casual) %>%
  summarize(total_rides = n(), .groups = 'drop') %>%
  arrange(year, month, member_casual)

print(summary)

```
Based on the table we can see that July has significantly more rides than January in both 2020 and 2024. Meaning more people want to ride during warmer season instead of cold season. We can see that there are much more member than casual riders in both month (Jan and July) and year (2020 and 2024). It looks like the useage of citi bike increase by years.

```{r}
#filter for July 2024 and summarize the top 5 starting stations
stations <- citi %>%
  filter(year == 2024, month == "July") %>%
  group_by(start_station_name) %>%
  summarize(total_rides = n(), .groups = 'drop') %>%
  arrange(desc(total_rides)) %>%
  head(5)


print(stations)

```

```{r}
#compute the median ride duration
duration <- citi %>%
  group_by(year, month, weekdays) %>%
  summarize(median_duration = median(duration, na.rm = TRUE), .groups = 'drop')

#produce plot
ggplot(citi, aes(x = weekdays, y = duration, fill = member_casual)) +
  geom_boxplot() +
  facet_grid(year ~ month) +
  labs(title = "Ride Duration by Day of Week, Month, and Year",
       x = "Day of Week",
       y = "Ride Duration (mins)",
       fill = "Membership Status") +
  theme_minimal()

```

The duration tend to be longer on weekends and members generally have shorter and more consistent ride durations throughout the week. July has longer ride duration compared to January across both years, a guess might be that people like to ride during warm day more. Ride during 2024 show more variability compare to 2020 and also be more longer than in 2020.

```{r}
#filter data for 2024 and produce plot
distribution <- citi %>%
  filter(year == 2024) %>%
  ggplot(aes(x = duration, fill = member_casual)) +
  geom_density(alpha = 0.6) +
  facet_grid(rideable_type ~ month) +
  labs(title = "Ride Duration Distribution in 2024 by Month, Membership, and Bike Type",
       x = "Ride Duration (mins)",
       y = "Density",
       fill = "Membership Status") +
  theme_minimal()

print(distribution)
```

By observing the plot, both classic and electric bikes is heavily right-skewed, meaning the majority of rides are relatively short, with ride duration concentrated under 30 minutes. However, electric bikes tend to have slightly shorter rides on average. The casual rider tend to have slightly longer rides compare to members. While member has shorter but more consistent ride duration. July has slightly broader distribution of duration compare to January,  especially for casual riders.

